### 2018-04-28 doc/paper.tex
Added a hacky introduction and some placeholder sections.
This introduction is largely modeled after and borrowed from
the two primary papers (K-fold Horner and compensated de Casteljau).

Also turned all `\[...\]` math blocks into `equation` groups so
that they get a number and can be referenced.

\renewcommand{\headrulewidth}{0pt}
\newcommand{\cond}[1]{\operatorname{cond}\left(#1\right)}
\newcommand{\fl}[1]{\operatorname{fl}\left(#1\right)}
\newcommand{\eps}{\varepsilon}

@@

\begin{abstract}
In computer aided geometric design a polynomial is usually represented in
Bernstein form. This paper presents a family of compensated algorithms to
accurately evaluate a polynomial in Bernstein form with floating point
coefficients. The principle is to apply error-free transformations to
improve the traditional de Casteljau algorithm. At each stage of computation,
round-off error is passed on to first order errors, then to second order
errors, and so on. After the computation has been ``filtered'' \(K\)-times
via this process, the resulting output is as accurate as the de Casteljau
algorithm performed in \(K\) times the working precision. Forward error
analysis and numerical experiments illustrate the accuracy of this family
of algorithms.
\end{abstract}

@@

@@

For a polynomial \(p(x)\) in the power basis, we have
(\cite{langlois_et_al:DSP:2006:442}):
\[\cond{p(x)} = \frac{\widetilde{p}\left(\left|x\right|\right)}{
  \left|p(x)\right|} = \frac{\sum_j \left|a_j\right| \left|x\right|^j}{
  \left|p(x)\right|}.\]
In particular, this means that if \(x \geq 0\) and each \(a_j \geq 0\)
we must necessarily have \(\cond{p(x)} = 1\). To see an example in use,
consider \(p(x) = (x - 1)^n\) and input values of the form \(x = 1 + \delta\)
(with \(\left|\delta\right| \ll 1\)). Since \(a_j = \binom{n}{j} (-1)^{n - j}\)
we have \(\widetilde{p}(x) = (x + 1)^n\) hence
\[\cond{p\left(1 + \delta\right)} = \frac{(2 + \delta)^n}{
  \left|\delta\right|^n} = \left|1 + \frac{2}{\delta}\right|^n.\]
As \(\delta \to 0\), this value approaches \(\infty\) (as expected).

For a polynomial \(p(s)\) in Bernstein form, we have (\cite{Jiang2010}):
\[\cond{p(s)} = \frac{\widetilde{p}\left(s\right)}{
  \left|p(s)\right|} = \frac{\sum_j \left|b_j\right| \left|B_{j, n}(s)\right|}{
  \left|p(s)\right|}.\]
The Bernstein form is suited for \(s \in \left[0, 1\right]\), which means
\(B_{j, n}(s) \geq 0\) typically. If \(s \in \left[0, 1\right]\) and each
\(b_j \geq 0\) we must necessarily have \(\cond{p(s)} = 1\). To see an
example in use, consider
\[p(s) = (1 - 2s)^n = \left[(1 - s) - s\right]^n = \sum_j \binom{n}{j}
(1 - s)^{n - j} (-s)^j = \sum_j (-1)^j B_{j, n}(s)\]
and input values of the form \(x = \frac{1}{2} + \delta\)
(with \(\left|\delta\right| \ll \frac{1}{2}\)). Since \(b_j = (-1)^j\)
we have \(\widetilde{p}(s) = \left[(1 - s) + s\right]^n = 1\)
\[\cond{p\left(\frac{1}{2} + \delta\right)} = \frac{1}{
  \left|2\delta\right|^n}.\]
As \(\delta \to 0\), this value approaches \(\infty\) (as expected).

@@

Consider
\[p(s) = (4s - 3)^3 (8s + 7) = (-189) B_{0, 4}(s) + (-54) B_{1, 4}(s) +
57 B_{2, 4}(s) + (-32) B_{3, 4}(s) + 15 B_{4, 4}(s)\]
at the point \(s = \frac{3}{4} + 800 u\). In exact arithmetic, we
have \(p(s) = 13(3200u)^3 + \mathcal{O}\left(u^4\right)\).
However, using the compensated de Casteljau algorithm results in
\(b_0^{(0)} + \partial b_0^{(0)} = 0\):

@@

\subsection{\texorpdfstring{\(K\)}{K}-Fold Error Filtering}

@@

After implementing for \(K = 2, 3, \ldots, 12\) and instrumenting all
relevant floating point operations, the \(K\)-fold Horner requires
\[(5 \cdot 2^K - 8)n + \left((K + 8) 2^K - 12K - 6\right) =
\mathcal{O}\left((n + K)2^K\right)\]
flops to evaluate a degree \(n\) polynomial (this only applies when
\(n \geq K - 1\)). As a comparison, the
non-compensated form of Horner requires \(2n\) flops. Of these,
\(\left(2^{K - 1} - 1\right)n - 2^{K - 1}(K - 3) - 2\) are
FMA (fused-multiply-add) instructions.

After implementing for \(K = 2, 3, 4, 5\) and instrumenting all relevant
floating point operations, the \(K\)-fold de Casteljau requires
\[(15K^2 - 34K + 26)T_n + K + 5 =
\mathcal{O}\left(n^2 K^2\right)\]
flops to evaluate a degree \(n\) polynomial. (Here \(T_n\) is the
\(n\)th triangular number.) As a comparison, the non-compensated form of
de Casteljau requires \(3 T_n + 1\) flops. Of these, \((3K - 4)T_n\) are
FMA instructions. On hardware that doesn't support FMA,
every FMA will be exchanged for 10 \(\ominus\)'s and 6 \(\otimes\)'s so the
count will increase by \((10 + 6 - 1)(3K - 4)T_n\).

>>>>

\renewcommand{\headrulewidth}{0pt}
\newcommand{\cond}[1]{\operatorname{cond}\left(#1\right)}
\newcommand{\fl}[1]{\operatorname{fl}\left(#1\right)}
\newcommand{\eps}{\varepsilon}
\newcommand{\mach}{\mathbf{u}}

@@

\begin{abstract}
In computer aided geometric design a polynomial is usually represented in
Bernstein form. This paper presents a family of compensated algorithms to
accurately evaluate a polynomial in Bernstein form with floating point
coefficients. The principle is to apply error-free transformations to
improve the traditional de Casteljau algorithm. At each stage of computation,
round-off error is passed on to first order errors, then to second order
errors, and so on. After the computation has been ``filtered'' \((K - 1)\)
times via this process, the resulting output is as accurate as the de Casteljau
algorithm performed in \(K\) times the working precision. Forward error
analysis and numerical experiments illustrate the accuracy of this family
of algorithms.
\end{abstract}

@@

\section{Introduction}

In computer aided geometric design, polynomials are usually expressed in
Bernstein form. Polynomials in this form are usually evaluated by the
de Casteljau algorithm. This algorithm has a round-off error bound
which grows only linearly with degree, even though the number of
arithmetic operations grows quadratically. Nevertheless the de Casteljau
algorithm returns results arbitrarily less accurate than the working
precision \(\mach\) when evaluating \(p(s)\) is ill-conditioned.
The relative accuracy of the computed
evaluation with the de Casteljau algorithm (\texttt{DeCasteljau}) satisfies
(\cite{Mainar1999}) the following a priori bound:
\begin{equation}\label{de-casteljau-error}
  \frac{\left|p(s) - \mathtt{DeCasteljau}(p, s)\right|}{\left|p(s)\right|} \leq
  \cond{p, s} \times \mathcal{O}(\mach).
\end{equation}
In the right-hand side of this inequality, \(\mach\) is the computing
precision and the condition number \(\cond{p, s} \geq 1\) only depends
on \(s\) and the Bernstein coefficients of \(p\) --- its expression will
be given further.

For ill-conditioned problems, such as evaluating \(p(s)\) near a
multiple root, the condition number may be arbitrarily large, i.e.
\(\cond{p, s} > 1 / \mach\), in
which case most or all of the computed digits will be incorrect.
In some cases, even the order of magnitude of the computed value
of \(p(s)\) can be incorrect.

To address ill-conditioned problems, error-free transformations can
be applied in \textit{compensated algorithms} to account for roundoff.
Error-free transformations were studied in great detail in \cite{Ogita2005}
and open a large number of applications.
In \cite{langlois_et_al:DSP:2006:442}, a compensated Horner's algorithm was
described to evaluate a polynomial in the monomial basis. In \cite{Jiang2010},
a similar method was described to perform a compensated version of the de
Casteljau algorithm. In both cases, the \(\cond{p, s}\) factor is moved
from \(\mach\) to \(\mach^2\) and the computed value is as accurate
as if the computations were done in twice the working precision. For example,
the compensated de Casteljau algorithm (\texttt{CompDeCasteljau}) satisfies
\begin{equation}\label{de-casteljau-2-error}
  \frac{\left|p(s) - \mathtt{CompDeCasteljau}(p, s)\right|}{
    \left|p(s)\right|} \leq \mach + \cond{p, s} \times
    \mathcal{O}\left(\mach^2\right).
\end{equation}
For problems with \(\cond{p, s} < 1 / \mach^2\), the relative error
is \(\mach\), i.e. accurate to full precision, aside from rounding to the
nearest floating point number.

In \cite{Graillat2009}, the authors built on the compensated Horner's
algorithm to produce a method for evaluating a polynomial as if
the computations were done in \(K\) times the working precision for
any \(K \geq 2\). This result motivates this paper, though the
approach there is somewhat different than ours. They perform each computation
with error-free transformations and interpret the errors as coefficients of new
polynomials. They then evaluate the error polynomials, which (recursively)
generate second order error polynomials and so on. This recursive property
causes the number of operations to grow exponentially in \(K\). Here, we
instead have a fixed number of error groups, each corresponding to roundoff
from the group above it. For example, when
\((1 - s) b_j^{(n)} + s b_{j + 1}^{(n)}\) is computed in floating point, any
error is filtered down to the error group below it.

As in~\eqref{de-casteljau-error}, the accuracy of the compensated
result~\eqref{de-casteljau-2-error} may be arbitrarily bad for ill-conditioned
polynomial evaluations. We describe how to defer rounding into progressively
smaller error groups and improve the accuracy of the computed result by a
factor of \(\mach\) for every error group added. So we derive
\texttt{CompDeCasteljauK}, a \(K\)-fold compensated de Casteljau algorithm
that satisfies the following a prior bound for any arbitrary integer \(K\):
\begin{equation}
  \frac{\left|p(s) - \mathtt{CompDeCasteljauK}(p, s)\right|}{
    \left|p(s)\right|} \leq \mach + \cond{p, s} \times
    \mathcal{O}\left(\mach^K\right).
\end{equation}
This means that the computed value with \texttt{CompDeCasteljauK} is now
as accurate as the result of the de Casteljau algorithm performed in
\(K\) times the working precision with a final rounding back to this
working precision.

The paper is organized as follows. Section~\ref{notation} introduces some
basic notations and results about floating point arithmetic, error-free
transformations and the de Casteljau algorithm. In Section~\ref{compensated-2},
the compensated algorithm for polynomial evaluation from \cite{Jiang2010} is
reviewed and notation is established for the expansion. In
Section~\ref{compensated-k}, the \(K\) compensated algorithm is provided and
a forward error analysis is performed. Finally, in Section~\ref{numerical} we
give numerical tests to illustrate the practical efficiency of our algorithms.

\section{Basic notation and results}\label{notation}

Placeholder

\section{Compensated de Casteljau}\label{compensated-2}

Placeholder

\section{\texorpdfstring{\(K\)}{K} Compensated de Casteljau}\label{compensated-k}

Placeholder

\section{Numerical experiments}\label{numerical}

Placeholder

@@

For a polynomial \(p(x)\) in the power basis, we have
(\cite{langlois_et_al:DSP:2006:442}):
\begin{equation}
\cond{p(x)} = \frac{\widetilde{p}\left(\left|x\right|\right)}{
  \left|p(x)\right|} = \frac{\sum_j \left|a_j\right| \left|x\right|^j}{
  \left|p(x)\right|}.
\end{equation}
In particular, this means that if \(x \geq 0\) and each \(a_j \geq 0\)
we must necessarily have \(\cond{p(x)} = 1\). To see an example in use,
consider \(p(x) = (x - 1)^n\) and input values of the form \(x = 1 + \delta\)
(with \(\left|\delta\right| \ll 1\)). Since \(a_j = \binom{n}{j} (-1)^{n - j}\)
we have \(\widetilde{p}(x) = (x + 1)^n\) hence
\begin{equation}
\cond{p\left(1 + \delta\right)} = \frac{(2 + \delta)^n}{
  \left|\delta\right|^n} = \left|1 + \frac{2}{\delta}\right|^n.
\end{equation}
As \(\delta \to 0\), this value approaches \(\infty\) (as expected).

For a polynomial \(p(s)\) in Bernstein form, we have (\cite{Jiang2010}):
\begin{equation}
\cond{p(s)} = \frac{\widetilde{p}\left(s\right)}{
  \left|p(s)\right|} = \frac{\sum_j \left|b_j\right| \left|B_{j, n}(s)\right|}{
  \left|p(s)\right|}.
\end{equation}
The Bernstein form is suited for \(s \in \left[0, 1\right]\), which means
\(B_{j, n}(s) \geq 0\) typically. If \(s \in \left[0, 1\right]\) and each
\(b_j \geq 0\) we must necessarily have \(\cond{p(s)} = 1\). To see an
example in use, consider
\begin{equation}
p(s) = (1 - 2s)^n = \left[(1 - s) - s\right]^n = \sum_j \binom{n}{j}
(1 - s)^{n - j} (-s)^j = \sum_j (-1)^j B_{j, n}(s)
\end{equation}
and input values of the form \(x = \frac{1}{2} + \delta\)
(with \(\left|\delta\right| \ll \frac{1}{2}\)). Since \(b_j = (-1)^j\)
we have \(\widetilde{p}(s) = \left[(1 - s) + s\right]^n = 1\)
\begin{equation}
\cond{p\left(\frac{1}{2} + \delta\right)} = \frac{1}{
  \left|2\delta\right|^n}.
\end{equation}
As \(\delta \to 0\), this value approaches \(\infty\) (as expected).

@@

Consider
\begin{equation}
p(s) = (4s - 3)^3 (8s + 7) = (-189) B_{0, 4}(s) + (-54) B_{1, 4}(s) +
57 B_{2, 4}(s) + (-32) B_{3, 4}(s) + 15 B_{4, 4}(s)
\end{equation}
at the point \(s = \frac{3}{4} + 800 u\). In exact arithmetic, we
have \(p(s) = 13(3200u)^3 + \mathcal{O}\left(u^4\right)\).
However, using the compensated de Casteljau algorithm results in
\(b_0^{(0)} + \partial b_0^{(0)} = 0\):

@@

\subsection{\texorpdfstring{\(K\)}{K} Error Filtering}

@@

After implementing for \(K = 2, 3, \ldots, 12\) and instrumenting all
relevant floating point operations, the \(K\)-fold Horner requires
\begin{equation}
(5 \cdot 2^K - 8)n + \left((K + 8) 2^K - 12K - 6\right) =
\mathcal{O}\left((n + K)2^K\right)
\end{equation}
flops to evaluate a degree \(n\) polynomial (this only applies when
\(n \geq K - 1\)). As a comparison, the
non-compensated form of Horner requires \(2n\) flops. Of these,
\(\left(2^{K - 1} - 1\right)n - 2^{K - 1}(K - 3) - 2\) are
FMA (fused-multiply-add) instructions.

After implementing for \(K = 2, 3, 4, 5\) and instrumenting all relevant
floating point operations, the \(K\)-fold de Casteljau requires
\begin{equation}
(15K^2 - 34K + 26)T_n + K + 5 =
\mathcal{O}\left(n^2 K^2\right)
\end{equation}
flops to evaluate a degree \(n\) polynomial. (Here \(T_n\) is the
\(n\)th triangular number.) As a comparison, the non-compensated form of
de Casteljau requires \(3 T_n + 1\) flops. Of these, \((3K - 4)T_n\) are
FMA instructions. On hardware that doesn't support FMA,
every FMA will be exchanged for 10 \(\ominus\)'s and 6 \(\otimes\)'s so the
count will increase by \((10 + 6 - 1)(3K - 4)T_n\).
